### 2021-05-23 project_ashish_monograph/CMS.tex
updated

\section{The CMS experiment}
\label{sec:cms}
(describe the experiment in general, its discovery, mention the international collaboration) \\

\onehalfspacing Compact Muon Solenoid (CMS) is one of the major experiment at CERN's Large Hadron Collider (LHC) that have been built to search for new physics. CMS is designed to detect new particles produced in the LHC's high energy proton-proton and heavy-ion (Lead, Xenon) collisions. CMS will also measure the properties (mass, decay width, coupling) of already discovered particles like quarks and leptons with unprecedented precision and will be looking to discover completely new, beyond standard model of particle physics phenomena which will help us to get a better understanding of our Universe and may confirm or rule out multiverse hypothesis \cite{Chatrchyan:1129810}. This experiment has already confirmed the discovery of the most important particle in SM, the Higgs boson which gives mass to all bosons and fermions via electroweak symmetry breaking mechanism \cite{Chatrchyan:2012xdj}. It has also detected the decay of Higgs boson to a pair of photons, W bosons, bottom quarks and muons.  \\

@@

The CMS detector is build around a large superconducting solenoid magnet and has a cylindrical shape with many concentric layers where various components of this detector are placed. It is placed at one of the collision points around the LHC ring and consists of Silicon tracker, Electromagnetic calorimeter, Hadron calorimeter and Muon chambers as shown in the left part of Fig. 4. \\

The large solenoid magnet bend charged particles and studying the path taken by these charged particles produced in proton-proton collisions using Silicon pixel detector (innermost part of CMS tracker) as they fly away from the interaction point, we can identify the charge of a particle as positive and negative charged particle tracks bend in opposite directions by the magnetic field and can also be used to calculate the momentum of the charged particle using curvature of the track as shown in the right part of Fig. 4. \\

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{./cmsdetector_merged.png}
  \caption{ \onehalfspacing Left: 3D design of CMS detector showing various concentric layers. Right: Transverse view of the CMS detector showing the silicon tracker, electromagnetic calorimeter, hadron calorimeter, superconducting solenoid and muon chambers. \cite{Chatrchyan:2008aa}}
  \label{fig:CMSdetector}
\end{figure}

@@

The pixel detector is the innermost of CMS tracker that contains 60 million pixels and these pixels assist in tracking the paths of particles emerging from the collision point. It is the nearest detector to the beam pipe and so will be vital in reconstructing the tracks of very short-lived particles. As it is very close to the collision point which means that the number of particles passing through is huge, the rate of particles received at 8cm from the beam line will be around 10 million particles per square centimetre per second. This feature is used to calculate the instantaneous luminosity $(L_{inst})$ by counting the number of pixel clusters which are collection of pixels hit by a track. The pixel detector is able to disentangle and reconstruct all the tracks left by the particles coming from the interaction point and has a long duration of 10 years. \\

The CMS Run 2 pixel detector consists of four concentric barrel layers (L1-L4) at radii of 29, 68, 109 and 160 mm and three disks (D1-D3) on each end at distances of 291, 396, and 516 mm from the center of the detector. It is built from 1856 segmented silicon sensor modules, where 1184 modules are used in the barrel pixel detector (BPIX) and 672 modules are used for the forward disks (FPIX) . \\

Each layer of the pixel detector is spilt into tiny square segments with each segment containing a silicon sensor having dimensions 100$\mu$m $\times$ 150$\mu$m as shown in the right part of Fig. 5. When a charged particle passes through this silicon sensor, it transfers energy to silicon atoms which causes electrons to be ejected from it creating electron-hole pairs. Each pixel uses an electric current to collect these charges on the pixel surface as a small electric signal. A electronic silicon chip is attached to each segment which amplifies the signal. The particle trajectory is deduced by knowing the hit pixels and as the pixel detector is made of two dimensional segments with four layers, we can create a three-dimensional picture of the path taken by the charged particle.

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./pixeldetector_merged1.png}
  \caption{ \onehalfspacing Left: Layout of the CMS Run 2 pixel detector compared to the original detector layout, in longitudinal view. Right: pixel detector diagram showing size of silicon sensor, readout chip and data buffers. \cite{TrackerGroupoftheCMS:2020bgg}}
  \label{fig:LHC}
\end{figure}

>>>>

\section{The CMS experiment}
\label{sec:cms}

\onehalfspacing Compact Muon Solenoid (CMS) is one of the major experiment at CERN's Large Hadron Collider (LHC) that have been built to search for new physics. CMS is designed to detect new particles produced in the LHC's high energy proton-proton and heavy-ion (Lead, Xenon) collisions. CMS will also measure the properties (mass, decay width, coupling) of already discovered particles like quarks and leptons with unprecedented precision and will be looking to discover completely new, beyond standard model of particle physics phenomena which will help us to get a better understanding of our Universe and may confirm or rule out multiverse hypothesis \cite{Chatrchyan:1129810}. This experiment has already confirmed the discovery of the most important particle in SM, the Higgs boson which gives mass to all bosons and fermions via electroweak symmetry breaking mechanism \cite{Chatrchyan:2012xdj}. It has also detected the decay of Higgs boson to a pair of photons, W and Z bosons, bottom quarks, muons and taus.  \\

@@

The CMS detector is build around a large superconducting solenoid magnet and has a cylindrical shape with many concentric layers where various components of this detector are placed. It is placed at one of the collision points around the LHC ring and consists of Silicon tracker, Electromagnetic calorimeter, Hadron calorimeter and Muon chambers as shown in the left part of Fig. 9. \\

The large solenoid magnet bend charged particles. Studying the path taken by these charged particles produced in proton-proton collisions using Silicon pixel detector (innermost part of CMS tracker) as they fly away from the interaction point, we can identify the charge of a particle as positive and negative charged particle tracks bend in opposite directions by the magnetic field. Particle trajectory can also be used to calculate the momentum of the charged particle using curvature of the track as shown in the right part of Fig. 9. \\

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{./cmsdetector_merged.png}
  \caption{ \onehalfspacing Left: 3D design of CMS detector showing various concentric layers. Right: Transverse view of the CMS detector showing the silicon tracker, electromagnetic calorimeter, hadron calorimeter, superconducting solenoid and muon chambers \cite{Chatrchyan:2008aa}.}
  \label{fig:CMSdetector}
\end{figure}

@@

The pixel detector is the innermost of CMS tracker that contains 60 million pixels and these pixels assist in tracking the paths of particles emerging from the collision point. It is the nearest detector to the beam pipe and so will be vital in reconstructing the tracks of very short-lived particles. As it is very close to the collision point which means that the number of particles passing through is huge, the rate of particles received at 8cm from the beam line will be around 10 million particles per square centimetre per second. This feature is used to calculate the instantaneous luminosity $(L_{inst})$ by counting the number of pixel clusters which are collection of pixels hit by a track.  \\

The CMS Run 2 pixel detector consists of four concentric barrel layers (L1-L4) at radii of 29, 68, 109 and 160 mm and three forward disks (D1-D3) on each end at distances of 291, 396, and 516 mm from the center of the detector as shown in the left part of Fig. 10. It is built from 1856 segmented silicon sensor modules, where 1184 modules are used in the barrel pixel detector (BPIX) and 672 modules are used for the forward disks (FPIX) as shown in the right part of Fig. 10. A module consists of one or two silicon sensors with 160 $\times$ 416 pixels which are small square units connected to 16 readout chips (ROCs). \\

Each layer of the pixel detector is spilt into tiny square segments with each segment called pixel containing a silicon sensor with dimensions 100$\mu$m $\times$ 150$\mu$m as shown in the right part of Fig. 10. When a charged particle passes through this silicon sensor, it transfers energy to silicon atoms which causes electrons to be ejected from it creating electron-hole pairs. Each pixel uses an electric current to collect these charges on the pixel surface as a small electric signal. A electronic silicon chip is attached to each segment which amplifies the signal. The particle trajectory is deduced by knowing the hit pixels and as the pixel detector is made of two dimensional segments with four layers, we can create a three-dimensional picture of the path taken by the charged particle. \\

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./pixeldetector_merged1.png}
  \caption{ \onehalfspacing Left: Layout of the CMS Run 2 pixel detector compared to the original detector layout, in longitudinal view. Right: pixel detector diagram showing size of silicon sensor, readout chip and data buffers \cite{TrackerGroupoftheCMS:2020bgg}.}
  \label{fig:LHC}
\end{figure}

\subsection{Cluster reconstruction}
Specific techniques are employed for the reconstruction of clusters in the pixel detector and all the reconstructed clusters are used for counting to calculate the instantaneous luminosity. Algorithm used for finding clusters check for pixels whose signal to noise ratio is more than 6 and then merge adjacent pixel. These algorithms estimates cluster position in two directions and cluster charge. A cluster is a set of adjacent pixels and only pixels above certain minimum value of charge are considered. Position of a cluster in X and Y directions are obtained by fitting X and Y projections to templates that are estimates for cluster shapes \cite{Chatrchyan:2014fea}. \\

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{./pixel_reco.png}
  \caption{ \onehalfspacing Diagram showing pixel cluster with charge accumulation in each pixel expressed in terms of thousands of electrons. Green numbers represent charge accumulation below the 2000 electron threshold which are excluded clusters. The dotted red line indicates the particle trajectory projection in the module plane and the red cross shows the position of true hit.}
  \label{fig:CMS}
\end{figure}

### 2021-05-23 project_ashish_monograph/LHC.tex
updated

\section{The Large Hadron Collider}
\label{sec:lhc}
(describe the accelerator complex, proton bunches, c.m. energy, all experiments, run periods, inst. and integrated lumi, Run 3 and HL-LHC plans) \\

\onehalfspacing CERN accelerator complex is a progression of machines namely Linac 2, Proton Synchrotron Booster (PSB), Proton Synchrotron (PS), Super Proton Synchrotron (SPS) and Large Hadron Collider (LHC) as shown in Fig. 1 that accelerate particles mainly protons to increasingly higher energies of 50 MeV, 1.4 GeV, 25 GeV, 450 GeV and 14 TeV respectively \cite {CERNBrochure2017002Eng}. Each machine is used to increase the energy of particle before the beam is injected into to the next machine. Linear accelerator (Linac2) is the starting point for protons used in various experiments at CERN. After proton energy reaches 50 MeV, protons enter PSB which further increases its energy before they are injected into PS which takes up its energy to 25 GeV. After reaching 25 GeV, protons enter SPS which further increases its energy to 450 GeV. The protons are finally transferred to two beam pipes of LHC kept at very high vacuum.  \\

The Large Hadron Collider (LHC) is the most powerful particle accelerator in the world primarily build to collide protons and heavy ions at 7, 8, 13 and 14 TeV during different LHC run period. The LHC consists of a 27 kilometre ring of superconducting magnets producing strong magnetic field which is used to guide particle beams along the LHC ring. Proton beam made up of bunches travelling in opposite directions kept in two ultrahigh vacuum tubes are accelerated close to the speed of light using time dependent electric field produced by radiofrequency cavities before they collide at four interaction points along the LHC ring where ATLAS, ALICE, CMS and LHCb detectors are placed \cite {CERNBrochure2017002Eng}. \\

The first proton beam was successfully maneuvered around the LHC ring in September 2008 that marked the beginning of Run 1 with proton beams colliding at 7-8 TeV center-of-mass energy and this run period ended in 2013 \cite{cmsrun} \cite{Alemany-Fernandez:1631030}. Run 2 started in 2015 with protons colliding at 13 TeV center-of-mass energy and continued till October 2018 \cite{Wenninger:2668326}. Run 3 is expected to begin at the start of March 2022 and High Luminosity (HL)-LHC phase is expected to begin in mid-2027 \cite{Dainese:2019rgk}. Proton Synchroton (PS) is responsible for providing proton bunches spaced 7 metres (25 ns) apart with each bunch containing more than 100 billion protons. The maximum number of bunches reached with the beam preparation method used during Run 2 is 2556 \cite{Wenninger:2668326}. The number of protons in the colliding bunches, time of separation between the bunches and effective area of the bunch which depend on the rms transverse bunch size along horizontal and vertical directions is one of the ways to calculate instantaneous luminosity (L_{inst}). \\

Luminosity is a key performance parameter of an accelerator as it is a measure of number of collisions. Luminosity can be categorized into two types: instantaneous and integrated luminosity. Instantaneous luminosity is defined as the number of collisions occuring per second in per centimeter square area while integrated luminosity is obtained by integrating instantaneous luminosity over a time interval. Both are very important variables as they are used to obtain the number of collision events. During Run 1, peak value of LHC instantaneous luminosity was $0.77 \times 10^{34} \: cm^{-2} s^{-1} $ with $30 \: fb^{-1}$ integrated luminosity \cite{cmsrun1lumi}. During Run 2, LHC instantaneous luminosity reached a peak value of $2.1 \times 10^{34}  \: cm^{-2} s^{-1}$ with $163 \: fb^{-1}$ integrated luminosity as shown in Fig. 2 \cite{CMS:2018elu}. At the end of Run 3, we expect to achieve an integrated luminosity  of $300 \:fb^{-1}$ while Phase II HL-LHC is expected to collect $3000 \:fb^{-1}$ of data as shown in Fig. 3 thereby increasing luminosity by a factor of 10 beyond the LHC's design value \cite{Dainese:2019rgk}. \\

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\columnwidth]{./LHCcomplex.png}
  \caption{\onehalfspacing Diagram of the LHC accelerator complex located in Geneva, Switzerland. Four interaction points are shown where the ALICE, ATLAS, LHCb and CMS detectors are stationed. \cite{Mobs:2684277}}
  \label{fig:LHC}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./peakintlumi_run12.png}
  \caption{\onehalfspacing Left: Instantaneous luminosity for Run 1 and Run 2 period. Right: Integrated luminosity for Run 1 and Run 2 period. \cite{lumi}}
  \label{fig:LHClumi}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7 \columnwidth]{./lumi_projection.png}
  \caption{ \onehalfspacing Projected performance of the LHC until 2038, which shows the preliminary dates for prolonged stops (LS1, LS2, LS3, LS4, LS5) of the LHC and luminosities. Red points show instantaneous luminosity $(L_{inst})$ while the blue line shows integrated luminosity. \cite{collaborations2019report}.}
  \label{fig:LHCPlans}
\end{figure}

>>>>

\section{The Large Hadron Collider}
\label{sec:lhc}

\subsection{CERN complex}

\onehalfspacing CERN accelerator complex is a progression of machines namely Linac 2, Proton Synchrotron Booster (PSB), Proton Synchrotron (PS), Super Proton Synchrotron (SPS) and Large Hadron Collider (LHC) as shown in Fig. 3 that accelerate particles mainly protons to increasingly higher energies of 50 MeV, 1.4 GeV, 25 GeV, 450 GeV and 14 TeV respectively \cite {CERNBrochure2017002Eng}. Each machine is used to increase the energy of particle before the beam is injected into to the next machine. Linear accelerator (Linac2) is the starting point for protons used in various experiments at CERN. After proton energy reaches 50 MeV, protons enter PSB which further increases its energy before they are injected into PS which takes up its energy to 25 GeV. After reaching 25 GeV, protons enter SPS which further increases its energy to 450 GeV. The protons are finally transferred to two beam pipes of LHC kept at very high vacuum to collide at the interaction points and then proton beam is dumped after sometime (1-35 hours). This entire procedure defines a single LHC fill generally consisting of $10^{14}$ protons, grouped into bunches to form the proton beam as shown in Fig. 4. \\

The Large Hadron Collider (LHC) is the most powerful particle accelerator in the world primarily build to collide protons and heavy ions at 7, 8, 13 and 14 TeV during different LHC run period. The LHC consists of a 27 km ring of superconducting magnets producing strong magnetic field which is used to guide particle beams along the LHC ring. Proton beam is made up of bunches travelling in opposite directions kept in two ultrahigh vacuum tubes are accelerated close to the speed of light using time dependent electric field produced by radiofrequency cavities before they collide at four interaction points along the LHC ring where ATLAS, ALICE, CMS and LHCb detectors are placed \cite {CERNBrochure2017002Eng}. \\

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\columnwidth]{./LHCcomplex.png}
  \caption{\onehalfspacing Diagram of the LHC accelerator complex located in Geneva, Switzerland. Four interaction points are shown where the ALICE, ATLAS, LHCb and CMS detectors are stationed \cite{Mobs:2684277}.}
  \label{fig:LHC}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./lhcfill.png}
  \caption{ \onehalfspacing A typical operational LHC cycle which involves set up (beam is accepted from injector complex), injection (beam is injected into LHC ring in form of bunch train), ramp (current in dipole increased to reach highest energy), squeeze (bunches are squeezed to have minimum proton separation in them), adjust (separation between beams at the interaction point is removed) and stable beams (collision settings are invoked and data taking period starts) \cite{fill}.}
  \label{fig:LHCPlans}
\end{figure}

\subsection {Run 1 and 2 of LHC}
The first proton beam was successfully maneuvered around the LHC ring in September 2008 that marked the beginning of Run 1 with proton beams colliding at 7-8 TeV center-of-mass energy and this run period ended in 2013 \cite{cmsrun} \cite{Alemany-Fernandez:1631030}. Run 2 started in 2015 with protons colliding at 13 TeV center-of-mass energy and continued till October 2018 \cite{Wenninger:2668326}. Run 3 is expected to begin at the start of March 2022 and High Luminosity (HL)-LHC phase is expected to begin in mid-2027 \cite{Dainese:2019rgk}. Proton Synchroton (PS) is responsible for providing proton bunches spaced 7 metres (25 ns) apart with each bunch containing more than 100 billion protons as shown in Fig. 5. The maximum number of bunches reached with the beam preparation method used during Run 2 is 2556 \cite{Wenninger:2668326}. The number of protons in the colliding bunches, time of separation between the bunches and effective area of the bunch which depend on the rms transverse bunch size along horizontal and vertical directions as shown in Fig. 6 is one of the ways to calculate instantaneous luminosity $(L_{inst})$. The number of proton proton collisions per bunch crossing is called pileup ($\mu$).\\

Rate of collision = $\sigma$ L = $(60 mb) (7.5 $\times$ 10^{34} cm^{-2} s^{-1})$ = $(60 $\times$ 10^{-3} $\times$ 10^{-24}) (7.5 $\times$ 10^{34})$ = 6000 million/s \\

where $1 \:barn = 10^{-24} cm^2$ and $1 \:mb = 10^{-27} cm^2$.\\

Colliding bunch rate = number of colliding bunches $\times$ LHC revolution frequency = 2800 $\times$ 11246\:Hz = 28.7\:MHz = 31.4 \:million/s \\

Pileup ($\mu$) = $\frac{Rate \:\:of \:\:collision}{colliding \:\:bunch\:\: rate} = \frac{6000 \:million/s}{31.4 \:million/s}$ = 200   for HL-LHC.    \\

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{./LHCReport_1_image_cut.png}
  \caption{ \onehalfspacing Diagram showing colliding bunches in each LHC beam separated by 25ns.}
  \label{fig:CMS}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\columnwidth]{./vdm_image.png}
  \caption{\onehalfspacing Figure showing two LHC bunches in green and blue approaching each other and colliding giving rise to bunch overlapping region shown in red.}
  \label{fig:CMS}
\end{figure}

\subsection{Luminosity and Phase II HL-LHC}
Luminosity is a key performance parameter of an accelerator as it is a measure of number of collisions. Luminosity can be categorized into two types: instantaneous and integrated luminosity. Instantaneous luminosity corresponds to the number of collisions occuring per second in per centimeter square area while integrated luminosity is obtained by integrating instantaneous luminosity over a time interval. Both are very important variables as they are used to obtain the number of collision events. During Run 1, peak value of LHC instantaneous luminosity was $0.77 \times 10^{34} \: cm^{-2} s^{-1} $ with $30 \: fb^{-1}$ integrated luminosity \cite{cmsrun1lumi}. During Run 2, LHC instantaneous luminosity reached a peak value of $2.1 \times 10^{34}  \: cm^{-2} s^{-1}$ with $163 \: fb^{-1}$ integrated luminosity as shown in Fig. 7 \cite{CMS:2018elu}. At the end of Run 3, we expect to achieve an integrated luminosity of $300 \:fb^{-1}$ while Phase II HL-LHC is expected to collect $3000 \:fb^{-1}$ of data as shown in Fig. 8 thereby increasing luminosity by a factor of 10 beyond the LHC's design value \cite{Dainese:2019rgk}. \\

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./peakintlumi_run12.png}
  \caption{\onehalfspacing Left: Instantaneous luminosity for Run 1 and Run 2 period. Right: Integrated luminosity for Run 1 and Run 2 period \cite{lumi}.}
  \label{fig:LHClumi}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7 \columnwidth]{./lumi_projection.png}
  \caption{ \onehalfspacing Projected performance of the LHC until 2038, which shows the preliminary dates for prolonged stops (LS1, LS2, LS3, LS4, LS5) of the LHC and luminosities. Red points show instantaneous luminosity $(L_{inst})$ while the blue line shows integrated luminosity \cite{collaborations2019report}.}
  \label{fig:LHCPlans}
\end{figure}

### 2021-05-23 project_ashish_monograph/PCC.tex
updated

\section{The Pixel Cluster Counting (PCC) method}
\label{sec:pcc}
(describe cluster reconstruction, the module selection/veto, cluster counting per bx, per LS,..) \\

Pixel cluster counting (PCC)  method is an offline technique for the calculation of instantaneous luminosity $(L_{inst})$ for any LHC run period by counting the number of pixel clusters in the pixel detector (innermost segment of the CMS tracker) in a zero bias event which requires that only two bunches cross at the CMS interaction point. The mean number of pixel clusters can be expressed in terms of mean number of pixel clusters per proton-proton interaction and the number of proton-proton interaction per bunch crossing that is also called pileup (PU). \\

$<N_{cluster}>$ = $<N_{cluster/interaction}>$ (PU)  or (PU) = $\frac{<N_{cluster}>}{<N_{cluster/interaction}>}$ \\

The instantaneous luminosity per bunch crossing $L_{inst/bunch}$ is proportional to the number of proton-proton (pp) interactions per bunch crossing (pileup) and the proportionality constant is the ratio of  LHC orbit frequency f and the pp interaction cross section $\sigma_{interaction}(\sqrt{s})$ where s is the proton-proton collision center-of-mass energy \cite{CMS-PAS-LUM-12-001}. \\

$L_{inst/bunch}$ = $\frac{f}{\sigma_{interaction}}$ (PU)  \\

$L_{inst/bunch}$ = $\frac{<N_{cluster}> \: f }{<N_{cluster/interaction}> \: \sigma_{interaction}(\sqrt{s})}$ \\

@@

$\sigma_{vis}$ = $<N_{cluster/interaction}>$  $\sigma_{interaction}(\sqrt{s})$ \\

\newpage Thus, instantaneous luminosity per bunch crossing is given by \\

$L_{inst/bunch}$ = $\frac{<N_{cluster}> \: \sigma_{interaction}(\sqrt{s}) \: f }{ \:\sigma_{vis} \: \sigma_{interaction}(\sqrt{s})}$ \\

$L_{inst/bunch}$ = $\frac{<N_{cluster}> \:\: f}{\sigma_{vis}}$ \\

The PCC visible cross section $\sigma_{vis}$ is determined using van der meer scan method which is described in section 5. Cluster counting can be done over different time periods like 1 bunch crossing (25ns), 1 Lumi section (23.36s), 4 Lumi nibble (1.46s) and similarly instantaneous luminosity can be calculated over these time periods using PCC method. \\

Specific techniques are employed for the reconstruction of clusters in the pixel detector and all the reconstructed clusters are used for counting to calculate the instantaneous luminosity. Algorithm used for finding clusters check for pixels whose signal to noise ratio is more than 6 and then merge adjacent pixel. These algorithms estimates cluster position in two directions and cluster charge. A cluster is a set of adjacent pixels and only pixels above certain minimum value of charge are considered. Position of a cluster in X and Y directions are obtained by fitting X and Y projections to templates that are estimates for cluster shapes \cite{Chatrchyan:2014fea}. \\

1. Run2018A \\

Run number ranges from 315255 to 316995 \\
Number of runs per period is 146 \\

2. Run2018B\\

Run number ranges from 317080 to 319311\\
Number of runs per period is 132\\

3. Run2018C\\

Run number ranges from 319337 to 320065\\
Number of runs per period is 88 \\

The veto lists corresponds to second Run2018D and select each detector part L2, L3, L4, D1, D2, D3 \cite{vetolist}. \\

1. vetolateRunDlowcuttight (Number of vetoed modules 1387) \\
2. vetolateRunDlowcuttightB1 (Number of vetoed modules 1850)\\
3. vetolateRunDlowcuttightB2 (Number of vetoed modules 1766)\\
4. vetolateRunDlowcuttightB3 (Number of vetoed modules 1641)\\
5. vetolateRunDlowcuttightF1 (Number of vetoed modules 1813)\\
6. vetolateRunDlowcuttightF2 (Number of vetoed modules 1799)\\
7. vetolateRunDlowcuttightF3 (Number of vetoed modules 1800)\\

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{./pixel_reco.png}
  \caption{ \onehalfspacing Diagram showing pixel cluster with charge accumulation in each pixel expressed in terms of thousands of electrons. Green numbers represent charge accumulation below the 2000 electron threshold which are excluded clusters. The dotted red line indicates the particle trajectory projection in the module plane and the red cross shows the position of true hit.}
  \label{fig:CMS}
\end{figure}

>>>>

\section{The Pixel Cluster Counting (PCC) method}
\label{sec:pcc}

\subsection{Luminosity determination based on PCC}
Pixel cluster counting (PCC)  method is an offline technique for the calculation of instantaneous luminosity $(L_{inst})$ for any LHC run period by counting the number of pixel clusters in the pixel detector (innermost segment of the CMS tracker) in a zero bias event which requires that only two bunches cross at the CMS interaction point. The mean number of pixel clusters $<N_{cl}>$ can be expressed in terms of mean number of pixel clusters per proton-proton interaction  $<N_{cl/pp}>$ and pileup ($\mu$) assuming that clusters do not overlap which is a reasonable assumption given the fine granularity of pixel detector. \\

$<N_{cl}> = <N_{cl/pp}> \mu$ \\

$\mu = \frac{<N_{cl}>}{<N_{cl/pp}>}$ \\

The instantaneous luminosity per bunch crossing $L_{b}$ is proportional to the number of proton-proton (pp) interactions per bunch crossing (pileup) and the proportionality constant is the ratio of  LHC orbit frequency f and the pp interaction cross section $\sigma_{pp}$ \cite{CMS-PAS-LUM-12-001}. \\

$L_{b}$ = $\frac{f}{\sigma_{pp}}$ $\mu$  or $L_{b}$ = $\frac{<N_{cl}> \: f }{<N_{cl/pp}> \: \sigma_{pp}}$ \\

@@

$\sigma_{vis}$ = $<N_{cl/pp}>$  $\sigma_{pp}$ \\

Thus, instantaneous luminosity per bunch crossing $L_b$ is given by \\

$L_{b}$ = $\frac{<N_{cl}> \: \sigma_{pp} \: f }{ \:\sigma_{vis} \: \sigma_{pp}}$ or $L_{b}$ = $\frac{<N_{cluster}> \:\: f}{\sigma_{vis}}$ \\

The PCC visible cross section $\sigma_{vis}$ is determined using van der Meer scan method which is described in section 5. Cluster counting can be done over different time periods like per bunch or orbit integrated, 1 Lumi section (23.36s), 4 Lumi nibble (1.46s) and similarly instantaneous luminosity can be calculated over these time periods using PCC method \cite{CMS:2018elu}. \\

\subsection{Run 2018 CMS Data} \\

CMS 2018 data is a collection of runs for each fill in different periods defined during the LHC operation period.

LHC 2018 run is divided into A, B, C and D as described below. \\
\begin{flushleft}
1. Run2018A \\
Run number ranges from 315255 to 316995 \\
Number of runs per period is 146 \\

2. Run2018B\\
Run number ranges from 317080 to 319311\\
Number of runs per period is 132\\

3. Run2018C \\
Run number ranges from 319337 to 320065\\
Number of runs per period is 88 \\

4.Run2018D \\
Run number ranges from 320500 to 323699 \\
Number of runs per period is 262 \\

5. late Run2018D \\
Run number ranges from 320500 to 325175 \\
Number of runs per period is 339 \\
\end{flushleft}

\subsection{Pixel detector Module Selection}
Selection of modules in pixel detector is required because module may experience dynamic inefficiency, bad module chips can stop sending data, chips may not work properly and need reconfiguring. Modules which are observed to function efficiently are considered for luminosity estimation. The module veto lists correspond to late Run2018D that select each detector part barrel layers L2, L3, L4 and forward disks D1, D2, D3 as shown in Fig. 12 and 13. Barrel layer 1 (L1) is not used for luminosity determination \cite{vetolist}. \\

\begin{flushleft}
1. Total number of vetoed modules 1387. \\
2. Number of vetoed modules for L2 1850.\\
3. Number of vetoed modules for L3 1766.\\
4. Number of vetoed modules for L4 1641.\\
5. Number of vetoed modules for D1 1813.\\
6. Number of vetoed modules for D2  1799.\\
7. Number of vetoed modules for D3 1800.\\
\end{flushleft}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8 \columnwidth]{./PDDisk_R12.png}
  \caption{ \onehalfspacing Diagram showing coordinates of modules in one forward disk of pixel detector having two rings. FPIX 0 correspond to ring 1 and FPIX 1 to ring 2. Yellow regions shows coordinates of vetoed modules while white regions shows coordinates of good modules used for luminosity determination.}
  \label{fig:CMS}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1 \columnwidth]{./bpixL2-L4.png}
  \caption{ \onehalfspacing Diagram showing z coordinate (signed module) and phi coordinate (signed ladder) of modules in three barrel layers (L2, L3 and L4) of pixel detector. Yellow regions shows coordinates of vetoed modules while white regions shows coordinates of good modules used in luminosity determination. }
  \label{fig:CMS}
\end{figure}

### 2021-05-23 project_ashish_monograph/bkg.tex
updated

\section{Backgrounds}
\label{sec:bkg}
(background in vdM, afterglow backgrounds in Physics, subtraction of these) \\

The cluster rate equation is modified in the presence of background events. \\

N = \sigma \:L_{int} + N_{background} \\

L_{int} = $ \int L_{inst} \: dt = \int \frac{R(t)}{\sigma_{vis}} \:dt $  \\

\sigma = \sigma_{interaction} (\sqrt{s}) \: A \:\epsilon\\

@@

$R(t) = \frac{d}{dt} <N_{cluster}>$ is the time dependent cluster rate. \\

$N_{background}$: Number of background events \\

$L_{inst}$: instantaneous luminosity.\\

$L_{int}$: integrated luminosity. \\

$\sigma_{vis}$: PCC visible cross section.\\

$\sigma_{interaction}$ : proton-proton interaction cross section. \\

s: center-of-mass energy of colliding protons. \\

A: acceptance of the pixel detector. \\

$\epsilon$ : efficiency of the pixel detector. \\

$N_{background}$ need to be subtracted during vdM scans for precise determination of $\sigma_{vis}$. During high pileup physics runs, background contribution to instantaneous luminosity/$\sigma_{vis}$ based on the pixel cluster counting (PCC) method arises from a small out-of-time response (OOTR, or afterglow) that has two types.\\

@@

Fig. 11 shows the afterglow noise induced by one colliding bunch. Since the distribution is normalized to 1 at the first bin it means the first bin is the colliding bunch. The values at bins $>$ 1 are the cluster counts observed for bunches after the colliding bunch divided by the cluster counts observed by the colliding bunch. There are three parts to this distribution: \\

@@

 Afterglow effects for various bunches in a beam is shown in Fig. 12 where the intial rising curve correspong to Type I and decay curve correspond to Type II afterglow effect. It was studied in data taken using random triggers (triggers that fire randomly in bunch crossings spread over the whole LHC orbit except for the abort gap), considering the pixel cluster counts in bunch crossings after bunch trains. The resulting correction depends on the LHC filling scheme and for a typical 2011 fill with 1380 bunches corresponds to a subtraction of close to 2.8$\%$ of the integrated luminosity per lumi section. The afterglow corrections for PCC during 2017 run are in the range 2-5 $\%$ for the Type I corrections and 2-3 $\%$ averaged over all active bunches for the Type II corrections \cite{Sirunyan:2759951}.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\columnwidth]{./PCC_background.png}
  \caption{ \onehalfspacing \cite{}.}
  \label{fig:CMS}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\columnwidth]{./SingleBunchAfterglow.png}
  \caption{Afterglow effect produced by a single colliding bunch.}
  \label{fig:LHC}
\end{figure}

>>>>

\section{Backgrounds}
\label{sec:bkg}
\subsection{vdM background}

The cluster rate equation is modified in the presence of background events.\\

N = \sigma \:L_{int} + N_{bg} \\

@@

$N_{bg}$: Number of background events \\

$N_{bg}$ need to be subtracted during vdM scans for precise determination of $\sigma_{vis}$. These contributions to the measured cluster rate are either subtracted before the fit as measured from noncolliding, unpaired bunches or from periods during which the beams were kept at a large distance, or determined in the fit as a constant term added to the fit function as shown in Fig. 13.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\columnwidth]{./PCC_bg.png}
  \caption{ \onehalfspacing Visible cross section $\sigma_{vis}$ with only background subtraction.}
  \label{fig:CMS}
\end{figure}

\subsection{High pileup physics run background}

During high pileup physics runs, background contribution to instantaneous luminosity based on the pixel cluster counting (PCC) method arises from a small out-of-time response (OOTR, or afterglow) that has two types.\\

@@

Fig. 14 shows the afterglow noise induced by one colliding bunch. Since the distribution is normalized to 1 at the first bin it means the first bin is the colliding bunch. The values at bins greater than 1 are the cluster counts observed for bunches after the colliding bunch divided by the cluster counts observed by the colliding bunch. There are three parts to this distribution: \\

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\columnwidth]{./SingleBunchAfterglow.png}
  \caption{Afterglow effect produced by a single colliding bunch.}
  \label{fig:LHC}
\end{figure}

 Afterglow effects for various bunches in a beam is shown in Fig. 15 where the intial rising curve correspond to Type I and decay curve correspond to Type II afterglow effect. It was studied in data taken using random triggers (triggers that fire randomly in bunch crossings spread over the whole LHC orbit except for the abort gap), considering the pixel cluster counts in bunch crossings after bunch trains. The resulting correction depends on the LHC filling scheme and for a typical 2011 fill with 1380 bunches corresponds to a subtraction of close to 2.8$\%$ of the integrated luminosity per lumi section. The afterglow corrections for PCC during 2017 run are in the range 2-5 $\%$ for the Type I corrections and 2-3 $\%$ averaged over all active bunches for the Type II corrections \cite{Sirunyan:2759951}.
